---
title: "bGEV model with weekly data"
author: "Alexandre Bouvier"
date: "`r Sys.Date()`"
output: html_document
---

# Tools

```{r}
library(INLA)
library(evd)
```

```{r}
giveme.gev.par = function(q, sbeta, alpha, beta, xi) 
{
  .mu = function(q, sbeta, alpha, beta, xi) {
    a = -log(1-beta/2)
    b = -log(beta/2)
    c = -log(alpha)
    if (all(xi > 0.0)) {
      tmp0 = (c^(-xi) - 1)/xi
      tmp1 = a^(-xi)
      tmp2 = b^(-xi)
      dbeta = (tmp1 - tmp2)/xi
      return(q - (sbeta/dbeta) * tmp0)
    } else if (all(xi == 0.0)) {
      dbeta = log(b) - log(a)
      tmp0 = log(c)
      return(q + (sbeta/dbeta) * tmp0)
    } else {
      stop("mixed case not implemented")
    }
  }
  
  .sigma = function(q, sbeta, alpha, beta, xi) {
    a = -log(1-beta/2)
    b = -log(beta/2)
    if (all(xi > 0.0)) {
      tmp1 = a^(-xi)
      tmp2 = b^(-xi)
      dbeta = (tmp1 - tmp2)/xi
      return(sbeta/dbeta)
    } else if (all(xi == 0.0)) {
      dbeta = log(b) - log(a)
      return(sbeta/dbeta)
    } else {
      stop("mixed case not implemented")
    }
  }
  
  return(list(mu = .mu(q, sbeta, alpha, beta, xi),
              sigma = .sigma(q, sbeta, alpha, beta, xi),
              xi = xi))
}
```

```{r}
map.tail = function(x, interval, inverse = FALSE) {
  if (!inverse) {
    return (interval[1] + (interval[2] - interval[1]) * exp(x)/(1.0 + exp(x)))
  } else {
    return (log((x-interval[1])/(interval[2]-x)))
  }
}
```


# bGEV model for maximum Ozone

## Load data
```{r}
data = read.csv("./weekly_ozone_and_covariates.csv")
summary(data)
```

```{r}
attach(data)
n = dim(data)[1]
```

##Define priors for the tail parameter 

The tail parameter of a GEV-type distribution must stay in an interval.
INLA uses an internal parametrisation, so we map an initial guess to internal scale.

```{r}
tail.interval = c(0, 0.5)
tail.intern = map.tail(0.1, tail.interval, inverse=TRUE)
```

```{r}
hyper.tail = list(initial = tail.intern, 
                  prior = "pc.gevtail",
                  param = c(7, tail.interval), 
                  fixed= FALSE)
```

##Prior for the scale (spread) parameter

```{r}
hyper.spread = list(prior = "loggamma",
                  param = c(3, 3),
                  initial = 1,
                  fixed = FALSE)
```


##Regression priors for spread covariates 

These are regression coefficients affecting  the spread of the bGEV.
```{r}
hyper.beta1 = hyper.beta2 = hyper.beta3 = hyper.beta4 = list(prior = "normal",
                  param = c(0, 1),
                  initial = 0)
```


##Assemble all bGEV hyperparameters

```{r}
hyper.bgev = list(
                  tail = hyper.tail,
                  spread = hyper.spread,
                  beta1 = hyper.beta1,
                  beta2 = hyper.beta2,
                  beta3 = hyper.beta3,
                  beta4 = hyper.beta4)
```


## Define quantile-based parametrisation
```{r}
p.alpha = 0.5 # median
p.beta = 0.5 # 50% interval
```

## Control list for the bGEV family

```{r}
control.bgev = list(q.location = p.alpha,
                    q.spread = p.beta,
                    # quantile levels for the mixing part
                    q.mix= c(0.05, 0.20), 
                    # the Beta(s1, s2) mixing distribution parameters. 
                    # Hard-coded for the moment: s1=s2=5
                    beta.ab = 5)
```

## scale the covariates

```{r}
data$minTemp   <- as.numeric(scale(data$minTemp))
data$maxTemp   <- as.numeric(scale(data$maxTemp))
data$windSpeed <- as.numeric(scale(data$windSpeed))
data$rainSum   <- as.numeric(scale(data$rainSum))

```


## Extract year and month

```{r} 
data$year = as.numeric(format(as.Date(data$week), "%Y")) - 2018
data$month = as.numeric(format(as.Date(data$week), "%m"))
```

## Train/test split

```{r}
train.mask = data$year + 2018 < 2024
station.filter = data$idSensor <= 9991

data.train = data[train.mask & station.filter,]
data.test = data[!train.mask & station.filter,]
```

## Tail covariates (empty)

we do not model covariates on the tail, so this is an empty matrix.

```{r}
tail.x = matrix(1, nrow(data.train), 1)
```

## Spread covariates & formula

```{r}

spread.x <- cbind(
  maxTemp_spread  = data.train$maxTemp,
  minTemp_spread  = data.train$minTemp,
  windSpeed_spread = data.train$windSpeed,
  rainSum_spread  = data.train$rainSum
)


formula = inla.mdata(maxOzone, spread.x, tail.x) ~ -1 + intercept + minTemp + maxTemp + windSpeed + rainSum

```


## build INLA input

```{r}
data.bgev = data.frame(data.train, 
                       intercept = 1,
                       spread.x = spread.x,
                       tail.x = tail.x)
```


##Fit the model
```{r}
fit.bgev.1 = inla(formula,
         family = "bgev",
         data = data.bgev,
         control.family = list(hyper = hyper.bgev,
                               control.bgev = control.bgev),
         control.fixed = list(prec=1),
         control.predictor = list(compute = TRUE),
        control.compute   = list(cpo = T, dic = F, waic = T, config = F),
        control.inla = list(int.strategy = "eb", strategy = "simplified.laplace"),
         verbose=FALSE)
```

## Save and reload the model

```{r}
saveRDS(fit.bgev.1, file = "bGEV_weekly_model_32_stations.rds")
```

```{r}
fit.bgev.1 = readRDS(file = "bGEV_weekly_model_32_stations.rds")
```


## Inspect fixed effects & hyperparameters

```{r}
round(fit.bgev.1$summary.fixed,4)

```


```{r}
round(fit.bgev.1$summary.hyperpar,4)

```

## Plot fixed effects with confidence intervals

```{r}
library(ggplot2)

fixed <- fit.bgev.1$summary.fixed
fixed$param <- rownames(fixed)

ggplot(fixed, aes(x = param, y = mean)) +
  geom_point() +
  geom_errorbar(aes(ymin = `0.025quant`, ymax = `0.975quant`)) +
  geom_hline(yintercept = 0, size=0.5, color="red") +
  ggtitle("location parameters heteroskedastic model") +
  coord_flip()
```


## PIT histogram (model checking)

```{r}
hist(fit.bgev.1$cpo$pit, breaks = 50, main = "PIT histogram", xlab = "PIT")

```

## Plot marginal posterior hyperparameters

```{r}
plot_marginal <- function(marg, main="") {
  plot(
    marg,
    type = "l",
    lwd = 2,
    xlab = "Parameter value",
    ylab = "Density",
    main = main
  )
}
par(mfrow=c(2,3))

plot_marginal(fit.bgev.1$marginals.hyperpar[["spread for BGEV observations"]], "Spread")
plot_marginal(fit.bgev.1$marginals.hyperpar[["tail for BGEV observations"]], "Tail")

plot_marginal(fit.bgev.1$marginals.hyperpar[["beta1 (spread) for BGEV observations"]], "MaxTemp: effect on spread")
plot_marginal(fit.bgev.1$marginals.hyperpar[["beta2 (spread) for BGEV observations"]], "MinTemp: effect on spread")
plot_marginal(fit.bgev.1$marginals.hyperpar[["beta3 (spread) for BGEV observations"]], "WindSpeed: effect on spread")
plot_marginal(fit.bgev.1$marginals.hyperpar[["beta4 (spread) for BGEV observations"]], "RainSum: effect on spread")


```

## Predictions

```{r}


spread.x <- cbind(
  maxTemp_spread  = data.test$maxTemp,
  minTemp_spread  = data.test$minTemp,
  windSpeed_spread = data.test$windSpeed,
  rainSum_spread  = data.test$rainSum
)

tail.x = matrix(1, nrow(data.test), 1)


data.pred.bgev = data.frame(data.test, 
                       intercept = 1,
                       spread.x = spread.x,
                       tail.x = tail.x)
```



```{r}
pred.bgev = inla(
  formula,
  family = "bgev",
  data = data.pred.bgev,
  control.family    = list(hyper = hyper.bgev,
                               control.bgev = control.bgev),
  control.mode      = list(theta = fit.bgev.1$mode$theta, restart = FALSE),
  control.predictor = list(compute = TRUE),
  control.compute   = list(config = TRUE),  # required for samples
  control.inla      = list(strategy = "simplified.laplace", int.strategy = "eb"),
  verbose = F
)
```


```{r}
pred_mean = pred.bgev$summary.fitted.values$mean
pred_lower = pred.bgev$summary.fitted.values$`0.025quant`
pred_upper = pred.bgev$summary.fitted.values$`0.975quant`

pred.df.1 = data.frame(
  idSensor = data.test$idSensor,
  date  = as.Date(data.test$week),
  observed = data.test$maxOzone,
  pred_mean  = pred_mean,
  pred_lower = pred_lower,
  pred_upper = pred_upper
)


```


```{r}

ggplot(pred.df.1[pred.df.1$idSensor == 5707,], aes(x = date, group=idSensor)) +
  geom_line(aes(  y = observed, colour = "observed")) +
  geom_line(aes( y = pred_mean, colour = "predicted")) +
  geom_ribbon(aes(ymin = pred_lower, ymax = pred_upper), alpha = 0.2) +
  ggtitle("sensor 5707: Heteroskedastic model")
```


## --- Predictive intervals via posterior simulation ---

Predictive intervals quantify the full uncertainty of future observations by combining parameter uncertainty and the inherent variability of the process.
Unlike confidence intervals on model parameters, they indicate the range in which new observations are expected to fall with a given probability.

```{r}
library(sn)

set.seed(123)
n_sims <- 1000

# Draw posterior samples of the latent predictor
samples <- inla.posterior.sample(
  n = n_sims,
  result = pred.bgev
)

# Extract linear predictors
eta_samples <- sapply(samples, function(s)
  s$latent[grep("^Predictor", rownames(s$latent))]
)

# Posterior means of hyperparameters
xi_hat    <- fit.bgev.1$summary.hyperpar["tail for BGEV observations","mean"]
sigma_hat <- fit.bgev.1$summary.hyperpar["spread for BGEV observations","mean"]

# Simulate observations
y_rep <- matrix(NA, nrow = n_sims, ncol = nrow(eta_samples))

for (i in 1:n_sims) {
  mu_i <- eta_samples[, i]
  y_rep[i, ] <- rgev(
    length(mu_i),
    loc   = mu_i,
    scale = sigma_hat,
    shape = xi_hat
  )
}

# Predictive intervals
pred_PI <- apply(y_rep, 2, quantile, probs = c(0.025, 0.5, 0.975))

pred.df$pred_PI_lower <- pred_PI[1, ]
pred.df$pred_PI_median <- pred_PI[2, ]
pred.df$pred_PI_upper <- pred_PI[3, ]
```

```{r}
ggplot(pred.df[pred.df$idSensor == 5707,], aes(x = date)) +
  geom_line(aes(y = observed, colour = "observed") )+
  geom_line(aes(y = pred_PI_median, colour = "predicted") )+
  geom_ribbon(
    aes(ymin = pred_PI_lower, ymax = pred_PI_upper),
    fill = "blue", alpha = 0.3
  ) +
  ggtitle("Predictive intervals (sensor 5707)")
```
The predicted mean captures the seasonal ozone pattern well. Predictive intervals are wide, which is expected for extreme-value models and reflect natural variability. Most observations lie within the intervals, indicating good overall predictive calibration.

## --- GEV Probability Integral Transform ---

Q–Q (PIT) plots assess how well the model’s predicted distribution matches the observed data.
Points close to the diagonal indicate good calibration, while systematic deviations reveal under- or overestimation in certain parts of the distribution.

```{r}
mu_hat <- pred.bgev$summary.fitted.values$mean
xi_hat <- fit.bgev.1$summary.hyperpar["tail for BGEV observations","mean"]
sigma_hat <- fit.bgev.1$summary.hyperpar["spread for BGEV observations","mean"]

pit_gev <- pgev(
  pred.df$observed,
  loc   = mu_hat,
  scale = sigma_hat,
  shape = xi_hat
)

qqplot(
  qunif(ppoints(length(pit_gev))),
  sort(pit_gev),
  main = "GEV Q–Q plot (PIT)",
  xlab = "Theoretical quantiles",
  ylab = "Empirical PIT"
)
abline(0, 1, col = "red")
```
Center: points are close to the diagonal
→ the model is well calibrated for typical values.

Lower probabilities (left tail): points lie above the diagonal
→ the model slightly overestimates low values (lower tail too heavy).

Higher probabilities (right tail): points lie below the diagonal
→ the model underestimates extreme high values (upper tail too light).

## --- Extreme quantile coverage ---

Coverage at a given quantile (e.g., 95%) measures the proportion of observed values below that predictive quantile.
Values close to the nominal probability indicate well-calibrated extremes, while values too high or low reveal over- or underestimation.

```{r}
q95 <- qgev(
  0.95,
  loc   = mu_hat,
  scale = sigma_hat,
  shape = xi_hat
)

coverage_95 <- mean(pred.df$observed <= q95)

coverage_95
```
coverage_95 = 1 means that all observations fall below the model’s 95% predictive quantile.
This indicates an overly conservative model, with the upper tail (or dispersion) being overestimated.


# Temporal Model

## Set the tail prior

```{r}
tail = 0.08
tail.interval = c(0, 0.5)
tail.intern = map.tail(tail, tail.interval, inverse=TRUE)
```

## PC prior for the tail

```{r}
lambda = 7

hyper.tail = list(initial = tail.intern, 
                  prior = "pc.gevtail",
                  param = c(lambda, tail.interval), 
                  fixed= FALSE)
```

## Prior for the spread

```{r}
hyper.spread = list(prior = "loggamma",
                  param = c(3, 3),
                  initial = 1,
                  fixed = FALSE)
```


## Combine hyperparameters for the bGEV family

```{r}
hyper.bgev = list(
                  tail = hyper.tail,
                  spread = hyper.spread)
```


## bGEV model-specific quantile and mixing settings

```{r}

## bGEV hyperparameters 
p.alpha = 0.5  # alpha for q_alpha
p.beta  = 0.5 # beta for s_beta
pa  = 0.05 # Probability pa for the mixing region
pb  = 0.2  # Probability pb for the mixing region
c1  = 5    # parameters of the Beta weight function. Hard-coded for the moment: c1=c2=5
```


```{r}
control.bgev = list(q.location = p.alpha,
                    q.spread   = p.beta,
                    q.mix      = c(pa, pb), 
                    beta.ab    = c1)
```


## scale the covariates

```{r}
data$minTemp   <- as.numeric(scale(data$minTemp))
data$maxTemp   <- as.numeric(scale(data$maxTemp))
data$windSpeed <- as.numeric(scale(data$windSpeed))
data$rainSum   <- as.numeric(scale(data$rainSum))

```


## Extract year and month

```{r} 
data$year = as.numeric(format(as.Date(data$week), "%Y")) - 2018
data$month = as.numeric(format(as.Date(data$week), "%m"))
```

## Train/test split

```{r}
train_mask = data$year + 2018 < 2024
station.filter = data$idSensor <= 9991

data.train = data[train_mask & station.filter,]
data.test = data[!train_mask & station.filter,]
```

## Prior for monthly seasonal effect

```{r}
cyclic      = T
hyper.month = list(prec = list(initial = log(1/0.024), fixed = TRUE))
```

## INLA control options

```{r}

control.family    = list(hyper = hyper.bgev, control.bgev = control.bgev)
control.predictor = list(compute = TRUE)
control.compute   = list(cpo = T, dic = F, waic = T, config = F)
control.inla      = list(strategy = "simplified.laplace", int.strategy = "eb")
```

## build INLA input

```{r}
spread.x <- matrix(1, nrow(data.train), 1)
tail.x   <- matrix(1, nrow(data.train), 1)

data.bgev = data.frame(data.train, 
                       intercept = 1,
                       spread.x = spread.x,
                       tail.x = tail.x)
```

## Define the model formula

```{r}
formula = inla.mdata(maxOzone, spread.x, tail.x) ~ -1 +
  intercept +
  year +
  f(month,
   model = "rw2",
   cyclic = TRUE,
   hyper = hyper.month,
   scale.model = TRUE) +
  minTemp + maxTemp + windSpeed + rainSum

```

## Fit the model

```{r}
fit.bgev.2 = inla(formula,
               family = "bgev",
               data   = data.bgev,
               control.family    = control.family,
               control.predictor = control.predictor,
               control.compute   = control.compute,
               control.inla      = control.inla,
               verbose           = F) 
```

## Save and reload the model

```{r}
saveRDS(fit.bgev.2, file = "bGEV_temporal_weekly_model_32_stations.rds")
```

```{r}
fit.bgev.2 = readRDS(file = "bGEV_temporal_weekly_model_32_stations.rds")
```



## Summaries

```{r}
round(fit.bgev.2$summary.fixed, 4)
```

```{r}
round(fit.bgev.2$summary.hyperpar, 4)

```


```{r}
round(fit.bgev.2$summary.random$month,4)
```


## Plot fixed effect posterior means

```{r}
library(ggplot2)

fixed <- fit.bgev.2$summary.fixed
fixed$param <- rownames(fixed)

ggplot(fixed, aes(x = param, y = mean)) +
  geom_point() +
  geom_errorbar(aes(ymin = `0.025quant`, ymax = `0.975quant`)) +
  geom_hline(yintercept = 0, color="red", size=0.5) +
  coord_flip() +
  ggtitle("parameters of the model")
```

## PIT histogram

```{r}
hist(fit.bgev.2$cpo$pit, breaks = 50, main = "PIT histogram", xlab = "PIT")

```

## Posterior marginals for spread and tail

```{r}
plot_marginal <- function(marg, main="") {
  plot(
    marg,
    type = "l",
    lwd = 2,
    xlab = "Parameter value",
    ylab = "Density",
    main = main
  )
}
par(mfrow=c(2,1))

plot_marginal(fit.bgev$marginals.hyperpar[["spread for BGEV observations"]], "Spread σ")
plot_marginal(fit.bgev$marginals.hyperpar[["tail for BGEV observations"]], "Tail ξ")


```

## Seasonal effect plot

```{r}
library(ggplot2)

month_effect <- fit.bgev.2$summary.random$month
month_effect$month <- month_effect$ID

ggplot(month_effect, aes(x = month, y = mean)) +
  geom_line() +
  geom_ribbon(aes(ymin = `0.025quant`, ymax = `0.975quant`), alpha = .2) +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Seasonal effect of month",
       y = "Effect on maxOzone (latent scale)",
       x = "Month")
```


## Predictions

```{r}
spread.x <- matrix(1, nrow(data.test), 1)
tail.x   <- matrix(1, nrow(data.test), 1)

data.pred.bgev = data.frame(data.test, 
                       intercept = 1,
                       spread.x = spread.x,
                       tail.x = tail.x)


```

```{r}
formula = inla.mdata(maxOzone, spread.x, tail.x) ~ -1 + intercept + year + f(month, model = 'rw2', hyper = hyper.month, cyclic = cyclic, scale.model = TRUE) + minTemp + maxTemp + windSpeed + rainSum
```


```{r}
pred.bgev = inla(
  formula,
  family = "bgev",
  data = data.pred.bgev,
  control.family    = fit.bgev.2$control.family,
  control.mode      = list(theta = fit.bgev.2$mode$theta, restart = FALSE),
  control.predictor = list(compute = TRUE),
  control.compute   = list(config = TRUE),  # required for samples
  control.inla      = list(strategy = "simplified.laplace")
)
```


```{r}
pred_mean = pred.bgev$summary.fitted.values$mean
pred_lower = pred.bgev$summary.fitted.values$`0.025quant`
pred_upper = pred.bgev$summary.fitted.values$`0.975quant`

pred.df.2 = data.frame(
  idSensor = data.test$idSensor,
  date  = as.Date(data.test$week),
  observed = data.test$maxOzone,
  pred_mean  = pred_mean,
  pred_lower = pred_lower,
  pred_upper = pred_upper
)


```


```{r}

ggplot(pred.df.2[pred.df.2$idSensor == 5707,], aes(x = date, group=idSensor)) +
  geom_line(aes(  y = observed, colour = "observed")) +
  geom_line(aes( y = pred_mean, colour = "predicted")) +
  geom_ribbon(aes(ymin = pred_lower, ymax = pred_upper), alpha = 0.2) +
  ggtitle("sensor 5707: temporal model")
```


```{r}
ggplot(pred.df[pred.df$idSensor == 6582,], aes(x = date, group=idSensor)) +
  geom_line(aes(  y = observed, colour = "observed")) +
  geom_line(aes( y = pred_mean, colour = "predicted")) +
  geom_ribbon(aes(ymin = pred_lower, ymax = pred_upper), alpha = 0.2) +
  ggtitle("sensor 6582")

```

```{r}
ggplot(pred.df[pred.df$idSensor == 9856,], aes(x = date, group=idSensor)) +
  geom_line(aes(  y = observed, colour = "observed")) +
  geom_line(aes( y = pred_mean, colour = "predicted")) +
  geom_ribbon(aes(ymin = pred_lower, ymax = pred_upper), alpha = 0.2) +
  ggtitle("sensor 9856")

```



## --- Predictive intervals via posterior simulation ---

Predictive intervals quantify the full uncertainty of future observations by combining parameter uncertainty and the inherent variability of the process.
Unlike confidence intervals on model parameters, they indicate the range in which new observations are expected to fall with a given probability.

```{r}
library(sn)

set.seed(123)
n_sims <- 1000

# Draw posterior samples of the latent predictor
samples <- inla.posterior.sample(
  n = n_sims,
  result = pred.bgev
)

# Extract linear predictors
eta_samples <- sapply(samples, function(s)
  s$latent[grep("^Predictor", rownames(s$latent))]
)

# Posterior means of hyperparameters
xi_hat    <- fit.bgev.2$summary.hyperpar["tail for BGEV observations","mean"]
sigma_hat <- fit.bgev.2$summary.hyperpar["spread for BGEV observations","mean"]

# Simulate observations
y_rep <- matrix(NA, nrow = n_sims, ncol = nrow(eta_samples))

for (i in 1:n_sims) {
  mu_i <- eta_samples[, i]
  y_rep[i, ] <- rgev(
    length(mu_i),
    loc   = mu_i,
    scale = sigma_hat,
    shape = xi_hat
  )
}

# Predictive intervals
pred_PI <- apply(y_rep, 2, quantile, probs = c(0.025, 0.5, 0.975))

pred.df$pred_PI_lower <- pred_PI[1, ]
pred.df$pred_PI_median <- pred_PI[2, ]
pred.df$pred_PI_upper <- pred_PI[3, ]
```

```{r}
ggplot(pred.df[pred.df$idSensor == 5707,], aes(x = date)) +
  geom_line(aes(y = observed, colour = "observed") )+
  geom_line(aes(y = pred_PI_median, colour = "predicted") )+
  geom_ribbon(
    aes(ymin = pred_PI_lower, ymax = pred_PI_upper),
    fill = "blue", alpha = 0.3
  ) +
  ggtitle("Predictive intervals (sensor 5707)")
```
The predicted mean captures the seasonal ozone pattern well. Predictive intervals are wide, which is expected for extreme-value models and reflect natural variability. Most observations lie within the intervals, indicating good overall predictive calibration.

## --- GEV Probability Integral Transform ---

Q–Q (PIT) plots assess how well the model’s predicted distribution matches the observed data.
Points close to the diagonal indicate good calibration, while systematic deviations reveal under- or overestimation in certain parts of the distribution.

```{r}
mu_hat <- pred.bgev$summary.fitted.values$mean
xi_hat <- fit.bgev.2$summary.hyperpar["tail for BGEV observations","mean"]
sigma_hat <- fit.bgev.2$summary.hyperpar["spread for BGEV observations","mean"]

pit_gev <- pgev(
  pred.df$observed,
  loc   = mu_hat,
  scale = sigma_hat,
  shape = xi_hat
)

qqplot(
  qunif(ppoints(length(pit_gev))),
  sort(pit_gev),
  main = "GEV Q–Q plot (PIT)",
  xlab = "Theoretical quantiles",
  ylab = "Empirical PIT"
)
abline(0, 1, col = "red")
```
Center: points are close to the diagonal
→ the model is well calibrated for typical values.

Lower probabilities (left tail): points lie above the diagonal
→ the model slightly overestimates low values (lower tail too heavy).

Higher probabilities (right tail): points lie below the diagonal
→ the model underestimates extreme high values (upper tail too light).

## --- Extreme quantile coverage ---

Coverage at a given quantile (e.g., 95%) measures the proportion of observed values below that predictive quantile.
Values close to the nominal probability indicate well-calibrated extremes, while values too high or low reveal over- or underestimation.

```{r}
q95 <- qgev(
  0.95,
  loc   = mu_hat,
  scale = sigma_hat,
  shape = xi_hat
)

coverage_95 <- mean(pred.df$observed <= q95)

coverage_95
```
coverage_95 = 1 means that all observations fall below the model’s 95% predictive quantile.
This indicates an overly conservative model, with the upper tail (or dispersion) being overestimated.


# Model Comparaison

## WAIC

```{r}
fit.bgev.1$waic$waic

```
The second model with trend and seasonality has a better predictive fit.

## PIT

```{r}
pit1 <- fit.bgev.1$cpo$pit
pit2 <- fit.bgev.2$cpo$pit

pit1 <- pit1[is.finite(pit1)]
pit2 <- pit2[is.finite(pit2)]

par(mfrow = c(1,2))

hist(pit1, breaks = 20, freq = FALSE,
     main = "PIT – heteroskedastic model",
     xlab = "PIT")

hist(pit2, breaks = 20, freq = FALSE,
     main = "PIT – temporal model",
     xlab = "PIT")

```

```{r}
thr <- quantile(data.train$maxOzone, 0.95)

pit1_tail <- pit1[data.train$maxOzone > thr]
pit2_tail <- pit2[data.train$maxOzone > thr]

hist(pit1_tail, breaks = 10, freq = FALSE,
     main = "Tail PIT – Model 1")

hist(pit2_tail, breaks = 10, freq = FALSE,
     main = "Tail PIT – Model 2")

```



