---
title: "bGEV model with weekly data"
author: "Alexandre Bouvier"
date: "`r Sys.Date()`"
output: html_document
---

# Tools

```{r}
library(INLA)
library(evd)
```

```{r}
giveme.gev.par = function(q, sbeta, alpha, beta, xi) 
{
  .mu = function(q, sbeta, alpha, beta, xi) {
    a = -log(1-beta/2)
    b = -log(beta/2)
    c = -log(alpha)
    if (all(xi > 0.0)) {
      tmp0 = (c^(-xi) - 1)/xi
      tmp1 = a^(-xi)
      tmp2 = b^(-xi)
      dbeta = (tmp1 - tmp2)/xi
      return(q - (sbeta/dbeta) * tmp0)
    } else if (all(xi == 0.0)) {
      dbeta = log(b) - log(a)
      tmp0 = log(c)
      return(q + (sbeta/dbeta) * tmp0)
    } else {
      stop("mixed case not implemented")
    }
  }
  
  .sigma = function(q, sbeta, alpha, beta, xi) {
    a = -log(1-beta/2)
    b = -log(beta/2)
    if (all(xi > 0.0)) {
      tmp1 = a^(-xi)
      tmp2 = b^(-xi)
      dbeta = (tmp1 - tmp2)/xi
      return(sbeta/dbeta)
    } else if (all(xi == 0.0)) {
      dbeta = log(b) - log(a)
      return(sbeta/dbeta)
    } else {
      stop("mixed case not implemented")
    }
  }
  
  return(list(mu = .mu(q, sbeta, alpha, beta, xi),
              sigma = .sigma(q, sbeta, alpha, beta, xi),
              xi = xi))
}
```

```{r}
# mapping between internal parametrisation of tail
map.tail = function(x, interval, inverse = FALSE) {
  if (!inverse) {
    return (interval[1] + (interval[2] - interval[1]) * exp(x)/(1.0 + exp(x)))
  } else {
    return (log((x-interval[1])/(interval[2]-x)))
  }
}
```


# Data preprocessing

## Load data
```{r}
data = read.csv("./weekly_ozone_and_covariates.csv")
summary(data)
```

```{r}
attach(data)
n = dim(data)[1]
```


## scale the covariates

```{r}
data$minTemp   <- as.numeric(scale(data$minTemp))
data$maxTemp   <- as.numeric(scale(data$maxTemp))
data$windSpeed <- as.numeric(scale(data$windSpeed))
data$rainSum   <- as.numeric(scale(data$rainSum))

```


## Extract year and month

```{r} 
data$year = as.numeric(format(as.Date(data$week), "%Y")) - 2018
data$month = as.numeric(format(as.Date(data$week), "%m"))
```

## Train/test split

```{r}
train.mask = data$year + 2018 < 2024
station.filter = data$idSensor <= 9991

data.train = data[train.mask & station.filter,]
data.test = data[!train.mask & station.filter,]
```

# bGEV heteroskedastic model

##Define priors for the tail parameter 

The tail parameter of a GEV-type distribution must stay in an interval.
INLA uses an internal parametrisation, so we map an initial guess to internal scale.

```{r}
tail.interval = c(0, 0.5)
tail.intern = map.tail(0.1, tail.interval, inverse=TRUE)
```

```{r}
hyper.tail = list(initial = tail.intern, 
                  prior = "pc.gevtail",
                  param = c(7, tail.interval), 
                  fixed= FALSE)
```

##Prior for the scale (spread) parameter

```{r}
hyper.spread = list(prior = "loggamma",
                  param = c(3, 3),
                  initial = 1,
                  fixed = FALSE)
```


##Regression priors for spread covariates 

These are regression coefficients affecting  the spread of the bGEV.
```{r}
hyper.beta1 = hyper.beta2 = hyper.beta3 = hyper.beta4 = list(prior = "normal",
                  param = c(0, 1),
                  initial = 0)
```


##Assemble all bGEV hyperparameters

```{r}
hyper.bgev = list(
                  tail = hyper.tail,
                  spread = hyper.spread,
                  beta1 = hyper.beta1,
                  beta2 = hyper.beta2,
                  beta3 = hyper.beta3,
                  beta4 = hyper.beta4)
```


## Define quantile-based parametrisation
```{r}
p.alpha = 0.5 # median
p.beta = 0.5 # 50% interval
```

## Control list for the bGEV family

```{r}
control.bgev = list(q.location = p.alpha,
                    q.spread = p.beta,
                    # quantile levels for the mixing part
                    q.mix= c(0.05, 0.20), 
                    # the Beta(s1, s2) mixing distribution parameters. 
                    # Hard-coded for the moment: s1=s2=5
                    beta.ab = 5)
```


## Tail covariates (empty)

we do not model covariates on the tail, so this is an empty matrix.

```{r}
tail.x = matrix(1, nrow(data.train), 1)
```

## Spread covariates & formula

```{r}

spread.x <- cbind(
  maxTemp_spread  = data.train$maxTemp,
  minTemp_spread  = data.train$minTemp,
  windSpeed_spread = data.train$windSpeed,
  rainSum_spread  = data.train$rainSum
)


formula = inla.mdata(maxOzone, spread.x, tail.x) ~ -1 + intercept + minTemp + maxTemp + windSpeed + rainSum

```


## build INLA input

```{r}
data.bgev = data.frame(data.train, 
                       intercept = 1,
                       spread.x = spread.x,
                       tail.x = tail.x)
```


##Fit the model
```{r}
fit.bgev.1 = inla(formula,
         family = "bgev",
         data = data.bgev,
         control.family = list(hyper = hyper.bgev,
                               control.bgev = control.bgev),
         control.fixed = list(prec=1),
         control.predictor = list(compute = TRUE),
        control.compute   = list(cpo = T, dic = F, waic = T, config = F),
        control.inla = list(int.strategy = "eb", strategy = "simplified.laplace"),
         verbose=FALSE)
```

## Save and reload the model

```{r}
saveRDS(fit.bgev.1, file = "bGEV_weekly_model_32_stations.rds")
```

```{r}
fit.bgev.1 = readRDS(file = "bGEV_weekly_model_32_stations.rds")
```


## Inspect fixed effects & hyperparameters

```{r}
round(fit.bgev.1$summary.fixed,4)

```


```{r}
round(fit.bgev.1$summary.hyperpar,4)

```

## Plot fixed effects with confidence intervals

```{r}
library(ggplot2)

fixed <- fit.bgev.1$summary.fixed
fixed$param <- rownames(fixed)

ggplot(fixed, aes(x = param, y = mean)) +
  geom_point() +
  geom_errorbar(aes(ymin = `0.025quant`, ymax = `0.975quant`)) +
  geom_hline(yintercept = 0, size=0.5, color="red") +
  ggtitle("location parameters heteroskedastic model") +
  coord_flip()
```


## PIT histogram (model checking)

```{r}
hist(fit.bgev.1$cpo$pit, breaks = 50, main = "PIT histogram", xlab = "PIT")

```

## Plot marginal posterior hyperparameters

```{r}
plot_marginal <- function(marg, main="") {
  plot(
    marg,
    type = "l",
    lwd = 2,
    xlab = "Parameter value",
    ylab = "Density",
    main = main
  )
}
par(mfrow=c(2,3))

plot_marginal(fit.bgev.1$marginals.hyperpar[["spread for BGEV observations"]], "Spread")
plot_marginal(fit.bgev.1$marginals.hyperpar[["tail for BGEV observations"]], "Tail")

plot_marginal(fit.bgev.1$marginals.hyperpar[["beta1 (spread) for BGEV observations"]], "MaxTemp: effect on spread")
plot_marginal(fit.bgev.1$marginals.hyperpar[["beta2 (spread) for BGEV observations"]], "MinTemp: effect on spread")
plot_marginal(fit.bgev.1$marginals.hyperpar[["beta3 (spread) for BGEV observations"]], "WindSpeed: effect on spread")
plot_marginal(fit.bgev.1$marginals.hyperpar[["beta4 (spread) for BGEV observations"]], "RainSum: effect on spread")


```

## Predictions

```{r}


spread.x <- cbind(
  maxTemp_spread  = data.test$maxTemp,
  minTemp_spread  = data.test$minTemp,
  windSpeed_spread = data.test$windSpeed,
  rainSum_spread  = data.test$rainSum
)

tail.x = matrix(1, nrow(data.test), 1)


data.pred.bgev = data.frame(data.test, 
                       intercept = 1,
                       spread.x = spread.x,
                       tail.x = tail.x)
```



```{r}
pred.bgev = inla(
  formula,
  family = "bgev",
  data = data.pred.bgev,
  control.family    = list(hyper = hyper.bgev,
                               control.bgev = control.bgev),
  control.mode      = list(theta = fit.bgev.1$mode$theta, restart = FALSE),
  control.predictor = list(compute = TRUE),
  control.compute   = list(config = TRUE),  # required for samples
  control.inla      = list(strategy = "simplified.laplace", int.strategy = "eb"),
  verbose = F
)
```


```{r}
pred_mean = pred.bgev$summary.fitted.values$mean
pred_lower = pred.bgev$summary.fitted.values$`0.025quant`
pred_upper = pred.bgev$summary.fitted.values$`0.975quant`

pred.df.1 = data.frame(
  idSensor = data.test$idSensor,
  date  = as.Date(data.test$week),
  observed = data.test$maxOzone,
  pred_mean  = pred_mean,
  pred_lower = pred_lower,
  pred_upper = pred_upper
)


```


```{r}

ggplot(pred.df.1[pred.df.1$idSensor == 5707,], aes(x = date, group=idSensor)) +
  geom_line(aes(  y = observed, colour = "observed")) +
  geom_line(aes( y = pred_mean, colour = "predicted")) +
  geom_ribbon(aes(ymin = pred_lower, ymax = pred_upper), alpha = 0.2) +
  ggtitle("sensor 5707: Heteroskedastic model")
```

# Temporal bGEV Model

## Set the tail prior

```{r}
tail = 0.08
tail.interval = c(0, 0.5)
tail.intern = map.tail(tail, tail.interval, inverse=TRUE)
```

## PC prior for the tail

```{r}
lambda = 7

hyper.tail = list(initial = tail.intern, 
                  prior = "pc.gevtail",
                  param = c(lambda, tail.interval), 
                  fixed= FALSE)
```

## Prior for the spread

```{r}
hyper.spread = list(prior = "loggamma",
                  param = c(3, 3),
                  initial = 1,
                  fixed = FALSE)
```


## Combine hyperparameters for the bGEV family

```{r}
hyper.bgev = list(
                  tail = hyper.tail,
                  spread = hyper.spread)
```


## bGEV model-specific quantile and mixing settings

```{r}

## bGEV hyperparameters 
p.alpha = 0.5  # alpha for q_alpha
p.beta  = 0.5 # beta for s_beta
pa  = 0.05 # Probability pa for the mixing region
pb  = 0.2  # Probability pb for the mixing region
c1  = 5    # parameters of the Beta weight function. Hard-coded for the moment: c1=c2=5
```


```{r}
control.bgev = list(q.location = p.alpha,
                    q.spread   = p.beta,
                    q.mix      = c(pa, pb), 
                    beta.ab    = c1)
```


## Prior for monthly seasonal effect

```{r}
cyclic      = T
hyper.month = list(prec = list(initial = log(1/0.024), fixed = TRUE))
```

## INLA control options

```{r}

control.family    = list(hyper = hyper.bgev, control.bgev = control.bgev)
control.predictor = list(compute = TRUE)
control.compute   = list(cpo = T, dic = F, waic = T, config = F)
control.inla      = list(strategy = "simplified.laplace", int.strategy = "eb")
```

## build INLA input

```{r}
spread.x <- matrix(1, nrow(data.train), 1)
tail.x   <- matrix(1, nrow(data.train), 1)

data.bgev = data.frame(data.train, 
                       intercept = 1,
                       spread.x = spread.x,
                       tail.x = tail.x)
```

## Define the model formula

```{r}
formula = inla.mdata(maxOzone, spread.x, tail.x) ~ -1 +
  intercept +
  year +
  f(month,
   model = "rw2",
   cyclic = TRUE,
   hyper = hyper.month,
   scale.model = TRUE) +
  minTemp + maxTemp + windSpeed + rainSum

```

## Fit the model

```{r}
fit.bgev.2 = inla(formula,
               family = "bgev",
               data   = data.bgev,
               control.family    = control.family,
               control.predictor = control.predictor,
               control.compute   = control.compute,
               control.inla      = control.inla,
               verbose           = F) 
```

## Save and reload the model

```{r}
saveRDS(fit.bgev.2, file = "bGEV_temporal_weekly_model_32_stations.rds")
```

```{r}
fit.bgev.2 = readRDS(file = "bGEV_temporal_weekly_model_32_stations.rds")
```



## Summaries

```{r}
round(fit.bgev.2$summary.fixed, 4)
```

```{r}
round(fit.bgev.2$summary.hyperpar, 4)

```


```{r}
round(fit.bgev.2$summary.random$month,4)
```


## Plot fixed effect posterior means

```{r}
library(ggplot2)

fixed <- fit.bgev.2$summary.fixed
fixed$param <- rownames(fixed)

ggplot(fixed, aes(x = param, y = mean)) +
  geom_point() +
  geom_errorbar(aes(ymin = `0.025quant`, ymax = `0.975quant`)) +
  geom_hline(yintercept = 0, color="red", size=0.5) +
  coord_flip() +
  ggtitle("parameters of the model")
```

## PIT histogram

```{r}
hist(fit.bgev.2$cpo$pit, breaks = 50, main = "PIT histogram", xlab = "PIT")

```

## Posterior marginals for spread and tail

```{r}
plot_marginal <- function(marg, main="") {
  plot(
    marg,
    type = "l",
    lwd = 2,
    xlab = "Parameter value",
    ylab = "Density",
    main = main
  )
}
par(mfrow=c(2,1))

plot_marginal(fit.bgev$marginals.hyperpar[["spread for BGEV observations"]], "Spread σ")
plot_marginal(fit.bgev$marginals.hyperpar[["tail for BGEV observations"]], "Tail ξ")


```

## Seasonal effect plot

```{r}
library(ggplot2)

month_effect <- fit.bgev.2$summary.random$month
month_effect$month <- month_effect$ID

ggplot(month_effect, aes(x = month, y = mean)) +
  geom_line() +
  geom_ribbon(aes(ymin = `0.025quant`, ymax = `0.975quant`), alpha = .2) +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Seasonal effect of month",
       y = "Effect on maxOzone (latent scale)",
       x = "Month")
```


## Predictions

```{r}
spread.x <- matrix(1, nrow(data.test), 1)
tail.x   <- matrix(1, nrow(data.test), 1)

data.pred.bgev = data.frame(data.test, 
                       intercept = 1,
                       spread.x = spread.x,
                       tail.x = tail.x)


```

```{r}
formula = inla.mdata(maxOzone, spread.x, tail.x) ~ -1 + intercept + year + f(month, model = 'rw2', hyper = hyper.month, cyclic = cyclic, scale.model = TRUE) + minTemp + maxTemp + windSpeed + rainSum
```


```{r}
pred.bgev = inla(
  formula,
  family = "bgev",
  data = data.pred.bgev,
  control.family    = fit.bgev.2$control.family,
  control.mode      = list(theta = fit.bgev.2$mode$theta, restart = FALSE),
  control.predictor = list(compute = TRUE),
  control.compute   = list(config = TRUE),  # required for samples
  control.inla      = list(strategy = "simplified.laplace")
)
```


```{r}
pred_mean = pred.bgev$summary.fitted.values$mean
pred_lower = pred.bgev$summary.fitted.values$`0.025quant`
pred_upper = pred.bgev$summary.fitted.values$`0.975quant`

pred.df.2 = data.frame(
  idSensor = data.test$idSensor,
  date  = as.Date(data.test$week),
  observed = data.test$maxOzone,
  pred_mean  = pred_mean,
  pred_lower = pred_lower,
  pred_upper = pred_upper
)


```


```{r}

ggplot(pred.df.2[pred.df.2$idSensor == 5707,], aes(x = date, group=idSensor)) +
  geom_line(aes(  y = observed, colour = "observed")) +
  geom_line(aes( y = pred_mean, colour = "predicted")) +
  geom_ribbon(aes(ymin = pred_lower, ymax = pred_upper), alpha = 0.2) +
  ggtitle("sensor 5707: temporal model")
```



# Model Comparaison

## WAIC

```{r}
fit.bgev.1$waic$waic

```
The second model with trend and seasonality has a better predictive fit.

## PIT

```{r}
pit1 <- fit.bgev.1$cpo$pit
pit2 <- fit.bgev.2$cpo$pit

pit1 <- pit1[is.finite(pit1)]
pit2 <- pit2[is.finite(pit2)]

par(mfrow = c(1,2))

hist(pit1, breaks = 20, freq = FALSE,
     main = "PIT – heteroskedastic model",
     xlab = "PIT")

hist(pit2, breaks = 20, freq = FALSE,
     main = "PIT – temporal model",
     xlab = "PIT")

```





# Hybrid bGEV Model

## Set the tail prior

```{r}
tail = 0.08
tail.interval = c(0, 0.5)
tail.intern = map.tail(tail, tail.interval, inverse=TRUE)
```

## PC prior for the tail

```{r}
lambda = 7

hyper.tail = list(initial = tail.intern, 
                  prior = "pc.gevtail",
                  param = c(lambda, tail.interval), 
                  fixed= FALSE)
```

## Prior for the spread

```{r}
hyper.spread = list(prior = "loggamma",
                  param = c(3, 3),
                  initial = 1,
                  fixed = FALSE)
```

These are regression coefficients affecting  the spread of the bGEV.
```{r}
hyper.beta1 = hyper.beta2 = list(prior = "normal",
                  param = c(0, 1),
                  initial = 0)
```


##Assemble all bGEV hyperparameters

```{r}
hyper.bgev = list(
                  tail = hyper.tail,
                  spread = hyper.spread,
                  beta1 = hyper.beta1,
                  beta2 = hyper.beta2)

```

## bGEV model-specific quantile and mixing settings

```{r}

## bGEV hyperparameters 
p.alpha = 0.5  # alpha for q_alpha
p.beta  = 0.5 # beta for s_beta
pa  = 0.05 # Probability pa for the mixing region
pb  = 0.2  # Probability pb for the mixing region
c1  = 5    # parameters of the Beta weight function. Hard-coded for the moment: c1=c2=5
```


```{r}
control.bgev = list(q.location = p.alpha,
                    q.spread   = p.beta,
                    q.mix      = c(pa, pb), 
                    beta.ab    = c1)
```


## Prior for monthly seasonal effect

```{r}
cyclic      = T
hyper.month = list(prec = list(initial = log(1/0.024), fixed = TRUE))
```

## INLA control options

```{r}

control.family    = list(hyper = hyper.bgev, control.bgev = control.bgev)
control.predictor = list(compute = TRUE)
control.compute   = list(cpo = T, dic = F, waic = T, config = F)
control.inla      = list(strategy = "simplified.laplace", int.strategy = "eb")
```

## build INLA input

```{r}

spread.x <- cbind(
  windSpeed_spread = data.train$windSpeed,
  rainSum_spread  = data.train$rainSum
)

tail.x   <- matrix(1, nrow(data.train), 1)

data.bgev = data.frame(data.train, 
                       intercept = 1,
                       spread.x = spread.x,
                       tail.x = tail.x)
```

## Define the model formula

```{r}
formula = inla.mdata(maxOzone, spread.x, tail.x) ~ -1 +
  intercept +
  year +
  f(month,
   model = "rw2",
   cyclic = TRUE,
   hyper = hyper.month,
   scale.model = TRUE) +
  minTemp + maxTemp + windSpeed + rainSum

```

## Fit the model

```{r}
fit.bgev.hybrid = inla(formula,
               family = "bgev",
               data   = data.bgev,
               control.family    = control.family,
               control.predictor = control.predictor,
               control.compute   = control.compute,
               control.inla      = control.inla,
               verbose           = F) 
```

## Save and reload the model

```{r}
saveRDS(fit.bgev.hybrid, file = "bGEV_hybrid_weekly_model_32_stations.rds")
```

```{r}
fit.bgev.hybrid = readRDS(file = "bGEV_hybrid_weekly_model_32_stations.rds")
```

## Summaries

```{r}
round(fit.bgev.hybrid$summary.fixed, 4)
```

```{r}
round(fit.bgev.hybrid$summary.hyperpar, 4)

```


```{r}
round(fit.bgev.hybrid$summary.random$month,4)
```



## Plot fixed effect posterior means

```{r}
library(ggplot2)

fixed <- fit.bgev.hybrid$summary.fixed
fixed$param <- rownames(fixed)

ggplot(fixed, aes(x = param, y = mean)) +
  geom_point() +
  geom_errorbar(aes(ymin = `0.025quant`, ymax = `0.975quant`)) +
  geom_hline(yintercept = 0, color="red", size=0.5) +
  coord_flip() +
  ggtitle("parameters of the model")
```

## PIT histogram

```{r}
hist(fit.bgev.hybrid$cpo$pit, breaks = 50, main = "PIT histogram", xlab = "PIT")

```

## Posterior marginals for spread and tail

```{r}
plot_marginal <- function(marg, main="") {
  plot(
    marg,
    type = "l",
    lwd = 2,
    xlab = "Parameter value",
    ylab = "Density",
    main = main
  )
}
par(mfrow=c(2,2))

plot_marginal(fit.bgev.hybrid$marginals.hyperpar[["spread for BGEV observations"]], "Spread")
plot_marginal(fit.bgev.hybrid$marginals.hyperpar[["tail for BGEV observations"]], "Tail")

plot_marginal(fit.bgev.hybrid$marginals.hyperpar[["beta1 (spread) for BGEV observations"]], "WindSpeed: effect on spread")
plot_marginal(fit.bgev.hybrid$marginals.hyperpar[["beta2 (spread) for BGEV observations"]], "RainSum: effect on spread")

```

## Seasonal effect plot

```{r}
library(ggplot2)

month_effect <- fit.bgev.hybrid$summary.random$month
month_effect$month <- month_effect$ID

ggplot(month_effect, aes(x = month, y = mean)) +
  geom_line() +
  geom_ribbon(aes(ymin = `0.025quant`, ymax = `0.975quant`), alpha = .2) +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Seasonal effect of month",
       y = "Effect on maxOzone (latent scale)",
       x = "Month")
```


## Predictions

```{r}
spread.x <- cbind(
  windSpeed_spread = data.test$windSpeed,
  rainSum_spread  = data.test$rainSum
)

tail.x   <- matrix(1, nrow(data.test), 1)

data.pred.bgev = data.frame(data.test, 
                       intercept = 1,
                       spread.x = spread.x,
                       tail.x = tail.x)


```

```{r}
formula = inla.mdata(maxOzone, spread.x, tail.x) ~ -1 + intercept + year + f(month, model = 'rw2', hyper = hyper.month, cyclic = cyclic, scale.model = TRUE) + minTemp + maxTemp + windSpeed + rainSum
```


```{r}
pred.bgev = inla(
  formula,
  family = "bgev",
  data = data.pred.bgev,
  control.family    = fit.bgev.hybrid$control.family,
  control.mode      = list(theta = fit.bgev.hybrid$mode$theta, restart = FALSE),
  control.predictor = list(compute = TRUE),
  control.compute   = list(config = TRUE),  # required for samples
  control.inla      = list(strategy = "simplified.laplace")
)
```


```{r}
pred_mean = pred.bgev$summary.fitted.values$mean
pred_lower = pred.bgev$summary.fitted.values$`0.025quant`
pred_upper = pred.bgev$summary.fitted.values$`0.975quant`

pred.df.2 = data.frame(
  idSensor = data.test$idSensor,
  date  = as.Date(data.test$week),
  observed = data.test$maxOzone,
  pred_mean  = pred_mean,
  pred_lower = pred_lower,
  pred_upper = pred_upper
)


```


```{r}

ggplot(pred.df.2[pred.df.2$idSensor == 5707,], aes(x = date, group=idSensor)) +
  geom_line(aes(  y = observed, colour = "observed")) +
  geom_line(aes( y = pred_mean, colour = "predicted")) +
  geom_ribbon(aes(ymin = pred_lower, ymax = pred_upper), alpha = 0.2) +
  ggtitle("sensor 5707: temporal model")
```




# bGEV model for comparison with stan

```{r}
tail.interval = c(0, 0.5)
tail.intern = map.tail(0.1, tail.interval, inverse=TRUE)
```

```{r}
hyper.tail = list(initial = tail.intern, 
                  prior = "pc.gevtail",
                  param = c(7, tail.interval), 
                  fixed= FALSE)
```

##Prior for the scale (spread) parameter

```{r}
hyper.spread = list(prior = "loggamma",
                  param = c(3, 3),
                  initial = 1,
                  fixed = FALSE)
```


##Regression priors for spread covariates 

These are regression coefficients affecting  the spread of the bGEV.
```{r}
hyper.beta1 = hyper.beta2 = hyper.beta3 = hyper.beta4 = list(prior = "normal",
                  param = c(0, 1),
                  initial = 0)
```


##Assemble all bGEV hyperparameters

```{r}
hyper.bgev = list(
                  tail = hyper.tail,
                  spread = hyper.spread)
```


## Define quantile-based parametrisation
```{r}
p.alpha = 0.5 # median
p.beta = 0.5 # 50% interval
```

## Control list for the bGEV family

```{r}
control.bgev = list(q.location = p.alpha,
                    q.spread = p.beta,
                    # quantile levels for the mixing part
                    q.mix= c(0.05, 0.20), 
                    # the Beta(s1, s2) mixing distribution parameters. 
                    # Hard-coded for the moment: s1=s2=5
                    beta.ab = 5)
```



## Train/test split

```{r}
train.mask = data$year + 2018 < 2024
station.filter = data$idSensor %in% c(5707, 5710, 5717)

data.train = data[train.mask & station.filter,]
data.test = data[!train.mask & station.filter,]
```

## Tail covariates (empty)

we do not model covariates on the tail, so this is an empty matrix.

```{r}
tail.x = matrix(1, nrow(data.train), 1)
spread.x = matrix(1, nrow(data.train), 1)

```

## Spread covariates & formula

```{r}



formula = inla.mdata(maxOzone, spread.x, tail.x) ~ -1 + intercept + minTemp + maxTemp + windSpeed + rainSum

```


## build INLA input

```{r}
data.bgev = data.frame(data.train, 
                       intercept = 1,
                       spread.x = spread.x,
                       tail.x = tail.x)
```


##Fit the model
```{r}
fit.bgev.3 = inla(formula,
         family = "bgev",
         data = data.bgev,
         control.family = list(hyper = hyper.bgev,
                               control.bgev = control.bgev),
         control.fixed = list(prec=1),
         control.predictor = list(compute = TRUE),
         control.compute   = list(cpo = T, dic = F, waic = T, config = T),
         control.inla = list(int.strategy = "eb", strategy = "simplified.laplace"),
         verbose=FALSE)
```


```{r}
# Posterior marginals
marg_fixed <- fit.bgev.3$marginals.fixed

# Convert to summary
inla_summary <- lapply(marg_fixed, function(m) {
  data.frame(
    mean = inla.emarginal(identity, m),
    sd   = sqrt(inla.emarginal(function(x) x^2, m) -
                inla.emarginal(identity, m)^2),
    q025 = inla.qmarginal(0.025, m),
    q50  = inla.qmarginal(0.5, m),
    q975 = inla.qmarginal(0.975, m)
  )
})

inla_summary <- do.call(rbind, inla_summary)

```

```{r}
marg_hyper <- fit.bgev.3$marginals.hyperpar

inla_hyper <- lapply(marg_hyper, function(m) {
  data.frame(
    mean = inla.emarginal(identity, m),
    sd   = sqrt(inla.emarginal(function(x) x^2, m) -
                inla.emarginal(identity, m)^2),
    q025 = inla.qmarginal(0.025, m),
    q50  = inla.qmarginal(0.5, m),
    q975 = inla.qmarginal(0.975, m)
  )
})

inla_hyper <- do.call(rbind, inla_hyper)

```


```{r}
inla_samples <- inla.posterior.sample(
  n = 5000,
  result = fit.bgev.3
)
```

```{r}
param_hyper <- names(inla_samples[[1]]$hyperpar)

extract_draw <- function(s) {
  as.numeric(s$hyperpar)
}

inla_mat <- t(vapply(
  inla_samples,
  extract_draw,
  FUN.VALUE = numeric(length(param_hyper))
))

colnames(inla_mat) <- param_hyper

```



